Calendar-based graphics for visualising people's daily schedules

### 2017-07-19 main.Rmd
spelling
"proof reading by Stuart. Many thanks to Stuart.

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organises and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Facetting by the usual features such as day and month, was not sufficient to examine the behaviour. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format variations and options. It is implemented in the R package \pkg{sugrrants}.

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time} provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurred events).
2. Since the data are often collected at daily frequency or a time scale more frequent than daily, they typically involve a large number of observations spanning over a long time period.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need in comparing and contrasting between locations.

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distincitve across these 3 sensors. There is an eye-catching spike occurred to State Library, which is caused by the annual event--White Night--on 20th of Feburary."}
```

@@

A collection of data plots, organised in a familiar format, offer a useful and intuitive way to tell a richer story and reveal more complex relationships. @Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods to arrange the data plots of many types into a preserved geographical grid in the \pkg{geofacet} package. They both attempt to have the self-contained data plots in spatial context.

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using coloured blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented to the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics [@wilkinson2006grammar; @wickham2010layered].

```{r flinders-2016, fig.height = 6.5, fig.cap = "The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in the clearer manner."}
```

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organised into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging form 28-31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

Each cell forms a canvas on which to draw the data. Consider the canvas to have limits $[0, 1]$ horizontally and vertically. For the pedestrian sensor data, within each cell hour is plotted horizontally and count is plotted vertically. Each variable is scaled to have values between $[0,1]$, using the minimum and maximum of all the data values to be displayed assuming of fixed scales. Let $h$ be the scaled hour, and $c$ the scaled count.

@@

Note that for the vertical direction, the top left is the starting point of the grid, hence the subtractions, and resulting negative values to lay out the cells. Within each cell, the starting position is bottom left.

The algorithm can be relatively easily extended to layout from a single month to a few years, based on the period of time to be visualised. $(M, N)$ can be determined by the user using the arguments of \code{nrow} and \code{ncol}. If the one would like to visualise data spanning over three years, for example, $M = 12$ and $N = 3$ seem to be an appropriate choice when assessing the differences of a given month across the years.

We only illustrate that grids are laid out over the horizontal direction in the paper. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above when the argument \code{dir} is set to \code{"v"}. It particularly benefits those users who get accustomed to calendars of vertical organisation in some countries.

There are a few options provided for the \code{frame_calendar} function to adjust the display and they are shown as follows:

Assuming that tidy data [@wickham2014tidy] is the underlining data structure, the \code{x} takes a variable that will be mapped to the x axis and the \code{y} mapped to the y axis for the later plotting. In Figure \ref{fig:flinders-2016}, for example, the \code{x} is the variable specifying the time of the day, and the \code{y} is the variable suggesting the hourly counts. The \code{date} argument is given by the date variable that determines the correct order in the calendar layout. We shall describe some of the arguments that allow for different displays.

The algorithm described in Section \ref{sec:algorithm} is for the most common calendar layout--"monthly" calendar, and it can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as "weekly" and "daily" calendar respectively. Which layout to be used is controlled by the \code{calendar} argument. Due to the way of the arrangement, the weekly calendar puts more emphases on days of a week over days of a year, whereas the daily one serves as the opposite. The monthly format can be considered as the advanced twist of both weekly and daily. Temporal patterns lead to which format to be employed. The weekly calendar can be a nice attempt if the most variations can be characterised by days of a week. On the other hand, the absence of weekly effect but the presence of yearly effect can direct to the daily calendar. When both are present, the monthly calendar appears to be a better choice.

When \code{polar = TRUE}, the polar transformation is carried out for the data. Again the computation remain similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} is considered as the spiral display of Figure \ref{fig:flinders-2016}.

Section \ref{sec:algorithm} discusses the implementation applied to the fixed scale which is using the whole range of all the data values. The \code{scale} argument controls the scaling of the display. The fixed scale (\code{fixed}) is the default, meaning the data values in all positions to be scaled. The remaining options include free scale within each cell (\code{free}), cells derived from the same day of the week (\code{free_wday}), or cells from the same day of the month (\code{free_mday}). It utilises the comparisons of absolute or relative values, and the emphases of various types of variations.

Grouping the cells based on the same time period gives rise to the different scales. For example, the minimums and maximums obtained from each cell, that is every day $(i, j)$ together with $(m, n)$, result in local scale. The overall variation gives way to the individual shape. Figure \ref{fig:flinders-free} is an example of plotting line charts scaled locally. The daily variation is made more distinctive, compared to Figure \ref{fig:flinders-2016}.

```{r flinders-free, fig.height = 6.5, fig.cap = "Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive at the expense of the magnitude comparison loss."}
```

## Reference lines and labels

@@

@@

The comparison of one sensor to another adds additional insights to the dataset, which is commonly done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns occurred to both train stations--Flinders Street Station and Flagstaff station--are together driven by the commuters on the work days; however, the former is much outnumbered than the latter during the weekends and public holidays. It suggests that Flagstaff station limits its functionality as simply the public transport hub for day-to-day commuters, but various activities take place around the Flinders Street Station other than commuting. Because Flinders Street Station closely lies to South Bank and other places of attractions where the locals go for leisure and the tourists for travelling. It can be noted from Figure \ref{fig:overlay} that State Library follows the similar temporal trend as Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year Evening Firework, has barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and State Library on these days.

@@

Placing multiple series on the common scale makes the magnitudes comparable between them; the glyphs are yet small in size, resulting in the overlapping problem. To avoid the problem, the calendar layout can be embedded into a series of plots for the different sensors. Figure \ref{fig:facet} presents the idea of the graphs of calendar plots. By doing so, the comparability with respect to the magnitudes is sacrificed for the emphasis of the individual sensor.

@@

The \code{frame_calendar} function does not constrain itself to mapping only temporal variable to the \code{x}. Figure \ref{fig:scatterplot} shows the lag scatterplot, where the current hourly count is assigned to the \code{x} and the lagged hourly count to the \code{y} at Flinders Street Station, organised in the daily calendar. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, which suggests that an hour of a work day with many pedestrians is likely to be followed by the time of either more pedestrians or fewer. Quite the contrary, the relationship is so consistent on the non-work days. This is a clear sign of bimodalily on work days whereas unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The newly computed coordinates not only work with the simple geoms such as lines and points, but also boxplot and other complex geoms. Figure \ref{fig:boxplot} uses the loess smooth line superimposed atop the side-by-side boxplots as an example. It shows the distribution of hourly counts across all the 43 sensors. In general, bimodality features work days whereas unimodality features the rest of the days. The last week of December is just the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and hanging out for the firework in the New Year Evening.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organises and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Facetting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format variations and options. It is implemented in the R package \pkg{sugrrants}.

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides more details of the temporal patterns by facetting on day of the week. The sensor data, like many temporal datasets on human behaviours, lends itself to a number of exploratory data visualisation challenges:

1. Variations primarily result from multiple time scales including time of day, day of week, and day of year (such as public holiday and recurring events).
2. Since the data are often collected at sub-daily frequencies, they typically involve a large number of observations.
3. Measurements of a single type are made at multiple locations at a given time point, which creates the need for comparing and contrasting between locations.

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plot about the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distincitve across these 3 sensors. There is an eye-catching spike which occurred at State Library, caused by the annual event--White Night--on 20th of Feburary."}
```

@@

A collection of data plots, organised in a familiar format, offer a useful and intuitive way to understand complex data. 
@Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods in the \pkg{geofacet} package to arrange data plots into a grid, while preserving the geographical layout. They both attempt to enable one to see the spatial context of each self-contained data plot.

Alternatively, calendar-based graphics turn out to be a useful tool in unfolding human-related activities over time. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holidays, and presents other known seasons like school vacations; all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than day. Time of day, which serves as one of the most important aspects in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using coloured blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using the curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

We propose a new algorithm via linear algebra tools to go beyond the calendar-based heatmap. The approach is developed with these conditions in mind: (1) to make time of day present in addition to the existing temporal components such as day of week and day of year, (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout, (3) to enable an overlaying plot consisting of multiple time series. The proposed algorithm has been implemented in the \code{frame_calendar} function in the \pkg{sugrrants} package [@R-sugrrants] using \proglang{R} [@R-base].

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is an implementation of the grammar of graphics [@wilkinson2006grammar; @wickham2010layered].

```{r flinders-2016, fig.height = 6.5, fig.cap = "The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a 3 by 4 monthly grid represents all the traffic in 2016. The disparities between weekday and weekend along with public holiday are immediately apparent."}
```

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organised into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging from 28-31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

Each cell forms a canvas on which to draw the data. Consider the canvas to have limits $[0, 1]$ horizontally and vertically. For the pedestrian sensor data, within each cell, hour is plotted horizontally and count is plotted vertically. Each variable is scaled to have values in $[0,1]$, using the minimum and maximum of all the data values to be displayed, assuming fixed scales. Let $h$ be the scaled hour, and $c$ the scaled count.

@@

Note that for the vertical direction, the top left is the starting point of the grid which is why subtraction is performed. The resulting negative values place the cells down the vertical direction. Within each cell, the starting position is the bottom left.

@@

There are several options provided for the \code{frame_calendar} function to initialize and adjust the display of a calander plot:

Assuming that tidy data [@wickham2014tidy] is the underlying data structure, the parameter \code{x} takes a variable that will be mapped to the x axis and the parameter \code{y} mapped to the y axis for plot construction. In Figure \ref{fig:flinders-2016}, for example, the \code{x} is the variable specifying the time of the day, and the \code{y} is the variable representing the hourly counts. The \code{date} argument is given by the date variable that determines the correct order of the calendar layout.

The algorithm can be extended to display data from a single month up to a few years. The number of rows and columns in the layout can be specified with the arguments \code{nrow} and \code{ncol}. If the one would like to visualise data spanning over three years, for example, \code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid out horizontally. The vertical direction can be enabled by swapping $i$ and $j$ in equation \ref{eq:grid}, which occurs when the argument \code{dir} is set to \code{"v"}. This benefits users who are accustomed to calendars that are organised vertically, which is common in some countries. Next we shall describe some of the arguments that allow for alternate displays.

The algorithm described in Section \ref{sec:algorithm} is for the most common calendar layout, the "monthly" calendar, but it can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as the "weekly" and "daily" calendar, respectively. The layout to be used is controlled by the \code{calendar} argument. The weekly calendar puts more emphasis on days of a week over days of a year, whereas the daily calender does the opposite. The monthly calender can be considered as the advanced variant of both weekly and daily. Temporal patterns motivate which variant should be employed. The weekly calendar is appropriate if most of the variation can be characterised by days of the week. On the other hand, the daily calendar should be used when there is a yearly effect but not a weekly effect in the data. When both effects are present, the monthly calendar would be a better choice.

When \code{polar = TRUE}, the polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:flinders-polar} is considered as the spiral display of Figure \ref{fig:flinders-2016}.

Section \ref{sec:algorithm} discusses the implementation applied to a fixed scale, which uses the range over all data values. The \code{scale} argument controls the scaling of the display. The fixed scale (\code{fixed}) is the default, meaning that the data values in all positions are scaled. The remaining options include: free scale within each cell (\code{free}), cells derived from the same day of the week (\code{free_wday}), or cells from the same day of the month (\code{free_mday}). The scaling allows for the comparisons of absolute or relative values, and the emphasis of different temporal variations.

Grouping the cells based on the same time period gives rise to the different scales. For example, the minimums and maximums obtained from each cell produce a local scale. The overall variation gives way to the individual shape. Figure \ref{fig:flinders-free} is an example of plotting line charts scaled locally. The daily variation is more distinctive compared to Figure \ref{fig:flinders-2016}.

```{r flinders-free, fig.height = 6.5, fig.cap = "Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is more difficult to compare the size of peaks between days."}
```

@@

The comparison of one sensor to another adds additional insights to the dataset, which is commonly done with an overlaying plot such as Figure \ref{fig:overlay}. For instance, the dominant patterns that occurred at Flinders Street and Flagstaff train stations are driven by the commuters on work days; however, Flinders Street station has higher pedesterian counts during the weekends and public holidays. This suggests that Flagstaff station limits its functionality on non-work days, but various activities take place around the Flinders Street Station other than commuting. Figure \ref{fig:overlay} also shows that State Library follows a similar temporal trend as Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year's Eve has barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and State Library.

@@

To avoid the overlapping problem, the calendar layout can be embedded into a series of subplots for the different sensors. Figure \ref{fig:facet} presents the idea of facetting calendar plots. This allows comparing the overall structure between sensors, while emphasising indivdual sensor variation.

@@

The \code{frame_calendar} function does not constrain itself to mapping only the temporal variable to the \code{x} argument. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organised in the daily calendar layout. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, suggesting that an hour of a work day with many pedestrians is likely to be followed in the next hour with either greater or fewer numbers of pedestrians. However, this pattern is not present on non-work days. There is a bimodality on work days while there is unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The newly computed coordinates can work with more complicated plots, such as the boxplot. Figure \ref{fig:boxplot} uses the loess smooth line superimposed on side-by-side boxplots. It shows the distribution of hourly counts across all 43 sensors during December. In general, bimodality features on work days whereas unimodality features on the rest of the days. The last week of December is the holiday season: people are off work on the day before Christmas, go shopping on the Boxing day, and stay out for the fireworks on New Year's Eve.
